---
title: "Estimates of excess mortality in Austria during the first three years of the
  COVID-19 pandemic"
author: "Authors"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(targets)
library(dplyr)
library(ggplot2)
library(here)
library(ggdist)
library(patchwork)
library(lubridate)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here())

source(here("R/plots/model-timeseries-plots.R"))
```

```{r, warning=FALSE, message=FALSE}
report_years <- c(2020, 2021, 2022)

pop_age_levels <- 
  c("up to 4 years old",
    "5 to 9 years old",
    "10 to 14 years old", 
    "15 to 19 years old", 
    "20 to 24 years old", 
    "25 to 29 years old", 
    "30 to 34 years old", 
    "35 to 39 years old", 
    "40 to 44 years old", 
    "45 to 49 years old",
    "50 to 54 years old", 
    "55 to 59 years old", 
    "60 to 64 years old", 
    "65 to 69 years old", 
    "70 to 74 years old", 
    "75 to 79 years old", 
    "80 to 84 years old", 
    "85 to 89 years old", 
    "90 to 94 years old",
    "95 plus years old")

color_scale_values <- c("Excess Deaths", "Reported COVID Deaths (EMS)", "Reported COVID Deaths (Statistik Austria)")

excess_estimate <- tar_read("prediction_weekly_totals_model_main") 

yearly_excess_estimate <- excess_estimate |>
  mutate(year=year(week_end)) |>
  ungroup() |>
  group_by(year, .draw) |>
  summarise(excess = sum(excess),
            expected = sum(.prediction))

covid_deaths <- tar_read("covid_cases_deaths") |>
  ungroup() |>
  arrange(week_start) |>
  mutate(cum_covid_deaths = cumsum(covid_deaths)) |>
  select(week_start, covid_deaths, cum_covid_deaths)

yearly_covid_deaths <- covid_deaths |>
  ungroup() |>
  mutate(year = year(week_start + days(6))) |>
  group_by(year) |>
  summarise(covid_deaths = sum(covid_deaths))

yearly_covid_deaths_statistik_at <- data.frame(
  year = c(2021, 2022),
  covid_deaths = c(4257+3600+1192, 8055)
)

get_yearly_stats <- function(years_selected, ci_level, yearly_excess_estimate) {
  draws <- yearly_excess_estimate |>
    ungroup() |>
    filter(year %in% years_selected) |>
    group_by(.draw) |>
    mutate(excess = sum(excess))
  
  bayestestR::describe_posterior(draws$excess, ci=ci_level) |>
    as.data.frame() |>
    mutate(Median = round(Median),
           CI_low = round(CI_low),
           CI_high = round(CI_high)) 
}

format_yearly_stats <- function(years_selected, excess_estimate = yearly_excess_estimate) {
  ci_70 <- get_yearly_stats(years_selected, 0.7, excess_estimate)
  ci_90 <- get_yearly_stats(years_selected, 0.9, excess_estimate)
  
  glue::glue("{ci_90$Median} [70% CI: {ci_70$CI_low} - {ci_70$CI_high}, 90% CI: {ci_90$CI_low} - {ci_90$CI_high}]")
}
  
```

# Abstract


# Introduction

# Methods

## Data sources

Reported deaths by week, 5-year age-group, and sex were obtained from the open-data
repository of the official Austrian statistics office, Statistik Austria. 

## Model

We compared deaths reported in Austria during the first three years of the COVID-19 pandemic
from 2020 to 2022 to deaths expected from pre-pandemic trends and patterns in all-cause mortality. 


Excess deaths, $\delta_{y,w,a,s}$, in a given year, $y$, week, $w$, 5-year age-group, $a$, and sex $s$, 
were defined as the difference between reported deaths, $Y_{y,w,a,s}$ and
model estimated expected deaths $E_{y,w,a,s}$.

$$
\delta_{y,w,a,s} = Y_{y,w,a,s} - E_{y,w,a,s}
$$

To estimate the expected number of deaths in a given week we formulated an
age-structured Bayesian GAM model using R and brms. Weekly seasonality of 
expected deaths was modeled using a cyclic cubic spline with a knot for each week, $h(w)$. 
Long term trends in changes in all-cause mortality were accounted for by a thin plate spline 
with knots every 4 years, $g(y)$. 

The model uses a negative binomial distribution for the expected number of deaths
to allow for over dispersion in reported deaths, $E_{y,w,a,s} \sim \textrm{NegBinomial}(\mu_{y,w,a,s}, \theta)$. 

The predictors, $\mu_{y,w,s,x}$, were linked into the model using a log link function. 
The population, $N_{y,a,s}$, in each year, $y$, 5-year age group, $a$, and sex, $s$, 
was multiplied by a constant normalization factor $\gamma_{a,s}$. 
This normalization factor was used to improve the numerical stability of the model and
was calculated for each age group and sex by dividing the mean number of deaths 
in the pre-pandemic period $\bar{Y}_{a,s}$ by the mean population in this age group and sex 
in the pre-pandemic period $\bar{N}_{a,s}$.


$$
\log (\mu_{y,w,a,s}) = \beta_{a,s} \log{(\gamma_{a,s} N_{y,a,s})} + g(y) + h(w)
$$

The P-score is defined as the percentage difference between actual all-cause 
mortality and expected all-ause mortality. 

$$
P_{y,w,a,s} = \frac{\delta_{y,w,a,s}}{E_{y,w,a,s}}
$$
Aggregate results, total weekly excess mortality, were obtained by 
summing over samples drawn from the posterior distribution of the model. 
Bayesian credible intervals were calculated using the equal-tailed method.

## Code and data availability

All code is available on GitHub at 
[github.com/fvalka/covid19-austria-excess-mortality](https://github.com/fvalka/covid19-austria-excess-mortality).
The data sets are available as open-data at Statistik Austria, 
[https://data.statistik.gv.at/](https://data.statistik.gv.at/) and at the 
Austrian COVID-19 open data portal 
[https://www.data.gv.at/en/data/austrian-covid-19-open-data-information-portal/](https://www.data.gv.at/en/data/austrian-covid-19-open-data-information-portal/).


# Results

In total we estimate that `r format_yearly_stats(report_years)` excess deaths occurred 
in Austria during the first three years of the COVID-19 pandemic. 
With `r format_yearly_stats(2020)` excess deaths in 2020, `r format_yearly_stats(2021)` 
excess deaths in 2021, and `r format_yearly_stats(2022)` excess deaths in 2022.

```{r, excess-over-time, warning=FALSE, message=FALSE, out.width = '85%', fig.width=8.5, fig.height=6, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Model estimated cumulative and weekly deaths compared to reported COVID deaths."}


cumulative_excess_plot <- plot_model_timeseries(excess_estimate, cum_excess) +
  geom_line(data=covid_deaths, aes(y=cum_covid_deaths, color="Reported COVID Deaths (EMS)")) +
  ggtitle("Cumulative deaths")

weekly_excess_plot <- plot_model_timeseries(excess_estimate, excess) +
  geom_line(data=covid_deaths, aes(y=covid_deaths, color="Reported COVID Deaths (EMS)")) +
  ggtitle("Weekly deaths")


(cumulative_excess_plot /
  weekly_excess_plot)  + plot_layout(guides = 'collect') & theme(legend.position = 'bottom')
```

```{r, excess-yearly-totals, warning=FALSE, message=FALSE, out.width = '85%', fig.width=8.5, fig.height=3.5, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Posterior distributions of the yearly sums  of estimated excess mortality and P-scores. "}
yearly_excess_deaths_plot <- excess_estimate |>
  mutate(year=year(week_end)) |>
  ungroup() |>
  group_by(year, .draw) |>
  summarise(excess = sum(excess),
            expected = sum(.prediction)) |>
  ggplot(aes(y=excess, x=factor(year))) +
  geom_violin(draw_quantiles = c(0.5), aes(color="Excess Deaths")) +
  geom_point(data=yearly_covid_deaths, 
             aes(y=covid_deaths, color="Reported COVID Deaths (EMS)"),
             size=4, 
             alpha=0.7) +
  geom_point(data=yearly_covid_deaths_statistik_at, 
             aes(y=covid_deaths, color="Reported COVID Deaths (Statistik Austria)"), 
             size=4, 
             alpha=0.7) +
  scale_y_continuous(limits = c(0, NA)) +
  ggsci::scale_color_npg(limits = color_scale_values) +
  ggsci::scale_fill_npg(limits = color_scale_values) +
  labs(y="Excess Deaths", x="Year") +
  ggpubr::theme_pubr() +
  labs(color=NULL)

yearly_pscore_plot <- excess_estimate |>
  mutate(year=year(week_end)) |>
  ungroup() |>
  group_by(year, .draw) |>
  summarise(excess = sum(excess),
            expected = sum(.prediction)) |>
  mutate(p_score = (excess/expected)) |>
  ggplot(aes(y=p_score, x=factor(year))) +
  geom_violin(draw_quantiles = c(0.5)) +
  scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
  ggsci::scale_color_npg(limits = color_scale_values) +
  ggsci::scale_fill_npg(limits = color_scale_values) +
  labs(y="P-Score", x="Year") +
  ggpubr::theme_pubr()

(yearly_excess_deaths_plot | yearly_pscore_plot) + 
  plot_layout(guides = 'collect') & 
  theme(legend.position = 'bottom')
```

# Discussion

Austria experienced excess mortality in all years of the COVID-19 pandemic
with no decline in 2022. Reported COVID-19 deaths match the excess mortality
in 2020, are at the upper end or slightly higher than excess mortality in 2021
and in 2022 there is a large difference between COVID deaths reported in the
Austrian epidemiological surveillance system (EMS) and COVID deaths, with COVID 
as the primary or a secondary cause, taken from reported cause-of-death 
statistics from Statistik Austria. 

The main strength of our method is that our model considers both changes in demography,
including changes caused by excess mortality in each year of the pandemic 
affecting the population size in the following years, as well as long term 
trends in excess mortality and weekly seasonality. 

# References

# Supplement 

## Long term trends and seasonality

The model contains smoothing curves for weekly seasonality $h(w)$ and yearly 
trends $g(y)$, which are shown in the figure. The yearly trend was fitted 
until 2019 and projected up to 2022. 

```{r, supplement-seasonlity-and-trend, warning=FALSE, message=FALSE, out.width = '75%', fig.width=8, fig.height=3.5, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Supplemental figure: Model estimaed parameters for weekly seasonality cyclic cubic splines and yearly trend thin plate splines on the log-link scale."}
weekly_seasonality_plot <- ggplotify::as.ggplot(function() plot(brms::conditional_smooths(tar_read("model_main_weekly_age_structured"), 
                         smooths = "s(week, bs = \"cc\", k = 53)",
                         prob = 0.9), 
     ask = FALSE)) +
  ggtitle("Weekly seasonality on link scale: h(w)")

yearly_trend_plot <- ggplotify::as.ggplot(function() plot(brms::conditional_smooths(tar_read("model_main_weekly_age_structured"), 
                         smooths = "s(year, k = 5)",
                         prob = 0.9,
                         int_conditions = list(year = 2000:2022)),
     ask = FALSE)) +
  ggtitle("Yearly trend on link scale: g(y)")

weekly_seasonality_plot | yearly_trend_plot

```

## Estimated expected mortality rates by age group and sex

The model is based upon an expected rate of deaths for each age group and sex.
This estimated rate of expected deaths in the model is obtained from the 
estimates of the $\beta_{a,s}$ parameters multiplied with the population
normalization factor $\gamma_{a,s}$. 

These estimates are rates before the yearly trend and weekly seasonality are
applied to them. 

```{r, supplement-mortality-rates, warning=FALSE, message=FALSE, out.width = '75%', fig.width=8, fig.height=4, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Supplemental figure: Estimaed mortality rates by age groups and sex, before yearly trends and weekly seasonality are applied. "}
mortality_rates_fit <- brms::ranef(tar_read("model_main_weekly_age_structured"),
                                   probs=c(0.05, 1-0.05)) |>
  as.data.frame() |>
  tibble::rownames_to_column(var="group")

pop_norm_factor <- tar_read("pop_norm_factor") |>
  mutate(group = stringr::str_c(age, "_", sex))

mortality_rates_fit |> 
  left_join(pop_norm_factor) |>
  mutate(Median = age.sex.Estimate.mean_deaths_log * pop_norm_factor * 1e5,
         CI_low = age.sex.Q5.mean_deaths_log * pop_norm_factor * 1e5,
         CI_high = age.sex.Q95.mean_deaths_log * pop_norm_factor * 1e5) |>
  ggplot(aes(x=Median, y=age, color=sex)) +
  geom_pointinterval(aes(xmin = CI_low, x=Median, xmax=CI_high)) +
  scale_x_log10() +
  scale_y_discrete(limits=pop_age_levels) +
  annotation_logticks(sides = "b") +
  ggpubr::theme_pubr() +
  labs(x="Model estimated rate of expected deaths per 100 000",
       y="Age group",
       color="Sex")
  

```

## Sensitivity Analysis 

### Model bias in pre-pandemic period

To investigate if there is any residual bias in the model for the fitted 
pre-pandemic years (up until 2019) we estimated the cumulative excess deaths 
for this period.
Since we consider all patterns in all-cause mortality, including seasonality 
and pre-pandemic excess as the baseline this should lead to a cumulative excess
from the year 2000 to 2019 of approximately 0. 

```{r, supplement-sensitivity-model-bias, warning=FALSE, message=FALSE, out.width = '85%', fig.width=8.5, fig.height=4, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Supplemental figure: Estimate of excess mortality in the pre-pandemic and pandemic period combined to investigate residual bias in the model after parameter estimation. "}
prediction_weekly_totals_sensitivty_bias <- tar_read("prediction_weekly_totals_sensitivty_bias")

plot_model_timeseries(prediction_weekly_totals_sensitivty_bias, cum_excess) 

```

### Non age-structured model

We also implemented a non age structured model similar to previously published 
results [WHO excess mortality ref] to investigate the effect of the age structuring
in the model compared to simpler methods. 

This model was also implemented as a Bayesian GAM using a negative binomial
distribution and log link function. Two smoothing functions were used
one cubic cylic spline for weekly seasonality, $n(w)$, and a thin plate regression spline
for yearly trends, $m(y)$, which unlike the main model contain a knot for each year to 
account for combined changes in longer term all-cause mortality trends and shift
in demographics. 

The log linked estimator for this model, $\nu_{y,w}$, is defined as:


$$
\log (\nu_{y,w}) = m(y) + n(w)
$$

Since this model can for later years of the pandemic not consider the excess 
mortality which occurred during the earlier years of the pandemic we would expect 
the estimates based upon pre-pandemic trends to be more biased towards lower 
estimates for each consecutive year in the pandemic. 

```{r, supplement-no-age-structure, warning=FALSE, message=FALSE, out.width = '85%', fig.width=8.5, fig.height=4, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Supplemental figure: Estimated cumulative excess mortality from a non age structured model with weekly seasonality and yearly trend."}
prediction_weekly_totals_sensitivty_no_age_structure <- tar_read("prediction_weekly_totals_sensitivty_no_age_structure")

plot_model_timeseries(prediction_weekly_totals_sensitivty_no_age_structure, cum_excess) +
  geom_line(data=covid_deaths, aes(y=cum_covid_deaths, color="Reported COVID Deaths (EMS)")) +
  ggtitle("Non-age structured model, sensitivty analysis")
```

```{r, supplement-non-age-structured-excess, warning=FALSE, message=FALSE, out.width = '85%', fig.width=8.5, fig.height=3.5, fig.align = 'center', dpi=150, fig.cap="", fig.cap="Supplement figure: Model estimated yearly excess model for the age-structured main model and the non-age structured sensitivty model and difference between the estimated excess deaths in both models. "}
yearly_excess_estimate_sensitivity_no_age_structure <- prediction_weekly_totals_sensitivty_no_age_structure |>
  mutate(year=year(week_end)) |>
  ungroup() |>
  group_by(year, .draw) |>
  summarise(excess = sum(excess),
            expected = sum(.prediction))

plot_yearly_excess_difference_no_age_structure <- yearly_excess_estimate_sensitivity_no_age_structure |>
  rename(excess_sensitivty = excess,
         expected_sensitivty = expected) |>
  inner_join(yearly_excess_estimate) |>
  filter(year %in% report_years) |>
  mutate(difference = excess - excess_sensitivty) |>
  ggplot(aes(x=factor(year), y=difference)) +
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggpubr::theme_pubclean() +
  labs(x="Year", y="Excess deaths") +
  ggtitle("Difference between models")

plot_yearly_excess_no_age_structure <- yearly_excess_estimate_sensitivity_no_age_structure |> 
  mutate(model="non age-structured, sensitivty") |>
  rbind(yearly_excess_estimate |> 
          mutate(model = "age-structred, main")) |>
  filter(year %in% report_years) |>
  ggplot(aes(x=factor(year), y=excess, color=model)) +
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggpubr::theme_pubclean() +
  labs(x="Year", y="Excess deaths") +
  ggtitle("Estimated excess deaths")

(plot_yearly_excess_no_age_structure | plot_yearly_excess_difference_no_age_structure) + 
  plot_layout(guides = 'collect') & 
  theme(legend.position = 'bottom')

```

Using this non age-structured model we estimated a total of 
`r format_yearly_stats(report_years, yearly_excess_estimate_sensitivity_no_age_structure)`
excess deaths for the years 2020 to 2022 compared to 
`r format_yearly_stats(report_years)` excess deaths from the age-structured 
main model. 

### Backtesting for a pre-pandemic estimation 
